A general Nullstellensatz for generalised spaces

### 2019-04-15 slides-birmingham2019.tex
More notes

@@

  There is a notion of a \emph{model} of a geometric theory in
  a given topos. For instance, a ring in the usual sense is a model of the
  theory of rings in the topos~$\Set$. The structure sheaf of a scheme~$X$ is a
  model in the topos~$\Sh(X)$ of set-valued sheaves on~$X$.}

@@

  \jnote{1-}{In case the theory~$\TT$ is a Horn theory (for instance if it is
  an equational theory), the \emph{term algebra} (the set of terms in the
  empty context modulo provable equality) is a model of~$\TT$. While such
  models do enjoy some nice categorical properties, they are in general
  \emph{not} the generic model.}
\end{frame}

@@

\begin{frame}{Working internally to toposes}
  \justifying
  Let~$\C$ be a site. We recursively define
  \[
    U \models \varphi \quad
    \text{(``$\varphi$ holds on~$U$'')}
  \]
  for objects~$U \in \C$ and \squiggly{formulas}~$\varphi$.
  Write~``$\Sh(\C) \models \varphi$'' for~$1 \models \varphi$.
  \footnotesize
  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
  U \models \top &\Ll& \text{true} \\
  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ \text{the
  empty family is a covering of~$U$} \\
  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
  U \models \varphi \wedge \psi &\Ll&
  \text{$U \models \varphi$ and $U \models \psi$} \\
  U \models \varphi \vee \psi &\Ll&
  \hcancel{\text{$U \models \varphi$ or $U \models
  \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $(U_i \to U)_i$} \\
  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
  U \models \varphi \Rightarrow \psi &\Ll&
  \text{for all~$V \to U$: }
  \text{$V \models \varphi$ implies $V \models \psi$} \\
  U \models \forall s \? F\_ \varphi(s) &\Ll&
  \text{for all $V \to U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
  % U \models \forall F\_ \varphi(F) &\Ll&
  % \text{for all $V \to U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
  U \models \exists s \? F\_ \varphi(s) &\Ll&
  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  &&
  \text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$}
  %U \models \exists F\_ \varphi(F) &\Ll&
  %\hcancel{\text{there exists a sheaf $F_0$ on $U$ such that $U \models
  %\varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  %&&
  %\text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
  %&& \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
  \end{array} \]
\end{frame}

>>>>

% \begin{document}

@@

  There is a notion of a \emph{model} of a geometric theory in
  a given topos. For instance, a ring in the usual sense is a model of the
  theory of rings in the topos~$\Set$. The structure sheaf of a scheme~$X$ is a
  model in the topos~$\Sh(X)$ of set-valued sheaves on~$X$.

  With \emph{topos} we mean Grothendieck topos, and as metatheory we use a
  constructive but impredicative flavour of English (which could be formalised
  by what is supported by the internal language of elementary toposes with an NNO).
  However the Nullstellensatz presented later makes no use of the
  subobject classifier, hence the results can likely be generalised to hold in
  a predicative metatheory or to hold for arithmetic universes.}

@@

  \jnote{1}{In case the theory~$\TT$ is a Horn theory (for instance if it is
  an equational theory), the \emph{term algebra} (the set of terms in the
  empty context modulo provable equality) is a model of~$\TT$. While such
  models do enjoy some nice categorical properties, they are in general
  \emph{not} the generic model.

  In general, the generic model cannot be realised as a set-based model (with
  a set for each sort, a map for each function symbol and so on). Sets are too
  constant for this purpose; the flexibility of sheaves (``variable sets'') is
  required: The generic model lives in the topos of set-valued
  sheaves over~$\C_\TT$.

  A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
  equivalence class (modulo provable equivalence) of a geometric
  formula~$\theta$ such that~$\TT$ proves
  \begin{enumerate}
    \item ``$\theta$ is a relation on~$A \times B$'': $(\theta \seq{\vec x,
    \vec y} \varphi \wedge \psi)$
    \item ``$\theta$ is total'': $(\varphi \seq{\vec x} \exists \vec y\_
    \theta)$
    \item ``$\theta$ is single-valued'': $(\theta \wedge \theta[\vec y'/\vec y]
    \seq{\vec x,\vec y,\vec y'} \vec y = \vec y')$
  \end{enumerate}

  \vspace*{-0.6em}
  A family~$(\{ \vec x_i\_ \varphi_i \} \xrightarrow{[\theta_i]} \{ \vec y\_
  \psi \})_i$ is a covering iff~$\TT$ proves~$(\psi \seq{\vec y} \bigvee_i
  \exists \vec x_i\_ \theta_i)$.

  The slides experiment with using the symbol~``$\yon$'' for the Yoneda
  embedding~$\C_\TT \to \Set[\TT]$, as in \fixedhref{XXX}{Elements
  of~$(\infty,1)$-category theory} by Emily Riehl and Dominic Verity.}

  \jnote{2}{The special case that the generic model of a theory~$\TT$ can be realised as
  a model in~$\Set$ occurs iff~$\TT$ is Morita-equivalent to the
  empty theory, that is, iff~$\TT$ has exactly one model in any topos.

  The special case that there exists at least some conservative~$\TT$-model
  in~$\Set$ occurs iff~$\TT$ has a conservative geometric expansion to a theory
  which is Morita-equivalent to the empty theory.}
\end{frame}

@@

\begin{frame}{Working internally to toposes}
  \justifying
  Let~$\C$ be a site. We recursively define
  \[
    U \models \varphi \quad
    \text{(``$\varphi$ holds on~$U$'')}
  \]
  for objects~$U \in \C$ and \squiggly{formulas}~$\varphi$.
  Write~``$\Sh(\C) \models \varphi$'' for~$1 \models \varphi$.
  \footnotesize
  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
  U \models \top &\Ll& \text{true} \\
  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ \text{the
  empty family is a covering of~$U$} \\
  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
  U \models \varphi \wedge \psi &\Ll&
  \text{$U \models \varphi$ and $U \models \psi$} \\
  U \models \varphi \vee \psi &\Ll&
  \hcancel{\text{$U \models \varphi$ or $U \models
  \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $(U_i \to U)_i$} \\
  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
  U \models \varphi \Rightarrow \psi &\Ll&
  \text{for all~$V \to U$: }
  \text{$V \models \varphi$ implies $V \models \psi$} \\
  U \models \forall s \? F\_ \varphi(s) &\Ll&
  \text{for all $V \to U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
  % U \models \forall F\_ \varphi(F) &\Ll&
  % \text{for all $V \to U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
  U \models \exists s \? F\_ \varphi(s) &\Ll&
  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  &&
  \text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$}
  %U \models \exists F\_ \varphi(F) &\Ll&
  %\hcancel{\text{there exists a sheaf $F_0$ on $U$ such that $U \models
  %\varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  %&&
  %\text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
  %&& \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
  \end{array} \]

  \jnote{1-}{
    Using the internal language of a topos, we can import all$^\star$ of
    constructive impredicative mathematics into the internal universe of a
    topos. XXX
  }
\end{frame}
